---
title: "R_DM_2"
output: html_document
---

Library importation
regarder stargazer si le temps

```{r}
library(haven)
library(dplyr)
```


# Data visualisation

```{r}
file = "data/DM_Subject_2_Data.dta"
df <- read_dta(file)

head(df)
```

# Part 1


In this part, we looked at employed people, thus we only keep observations in df with a wage associated 

```{r}

df_wage <- df %>% 
  filter(!is.na(salhoraire))

```


`r nrow(df_wage)` observations over `r nrow(df)` have a mensual salary indicated. Indeed the wage is only available in the first and last quarter of the dataframe for employed individuals.



# Question 1 

We ignore the panel dimension in this question by keeping only the last observed quarter for employed individuals.

```{r}

df_wage_last <- df_wage %>% 
  filter(t == 6)

```



```{r}
experience = age - 18
dependent_variable <- "salhoraire"
control_variables <- c("origin", "homme", "experience", "experience_carré", "tuu_r")
as.factor --> pour valeur non ordonnale


```


















