---
title: "R_DM_2"
output: html_document
---

Library importation
regarder stargazer si le temps

```{r necessary library}
library(haven)
library(dplyr)
library(forcats)
library(stargazer)
```


# Data visualisation

```{r data importation}
file = "data/DM_Subject_2_Data.dta"
df <- read_dta(file)

# correction of the salmensuel's computing 
df$salhoraire <- df$salmensuel / (df$heuretra * 4.33)

head(df)
```

# Part 1


In this part, we looked at employed people, thus we only keep observations in df with a wage associated 

```{r filtering data by hourly wage}

df_wage <- df %>% 
  filter(!is.na(salhoraire))

```


`r nrow(df_wage)` observations over `r nrow(df)` have a mensual salary indicated. Indeed the wage is only available in the first and last quarter of the dataframe for employed individuals.



# Question 1 

We ignore the panel dimension in this question by keeping only the last observed quarter for employed individuals.

```{r filtering data by timestamp}

df_last_wage <- df_wage %>% 
  filter(t == 6)

```


Creation of new interest variables used in our linear regression.
Revised_age is a variable where we rescale age in order to have a quadratic relationship between the hourly salary and the age of the worker.

```{r creation of interest variables}

df_last_wage$revised_age <- df_last_wage$age - min(df_last_wage$age)

# capturing the quadratic relation
df_last_wage$revised_age_squared <- df_last_wage$revised_age ^ 2

```


## a) Linear regression results


For interest variables we obviously want to include origine because that's what interest us in the regression. To ensure capturing the real prediction of origine on the hourly wage, we need to include other variables that let unchanged the hourly salary all other things being equal. These variables should be related to the hourly wage but not with the origine. We thus selected 4 variables : 

  - "homme", an indicator of being a man (0 if a woman, 1 else)
  - "revised_age" and "revised_age_squared" (as described above)

These first two variables can quite likely be decorrelated of the origine of people

Lastly we hypothesize that the size of the urban area where people lived are not correlated to there origin which could be debate because rural area are often inhabted by french origin persons.

```{r interest variables}

# as_factor used for nominal variables
interest_variables <- c("factor(origine)", "factor(homme)", "revised_age", "revised_age_squared", "factor(tuu_r)") 

# as_factor takes as an entry a variable under character or factor type
df_last_wage$homme <- as.character(df_last_wage$homme)

# as_factor interpret coefficients relatively to the first level in alphebitc order. To have an easier interpretation of the coefficient we will ensure that the compared modality will be France (01) for origin, Man (1) for homme and rural municipality for tuu_r (1)

df_last_wage <- df_last_wage %>% 
  mutate(
    # 01 corresponds to a french origin
    origine <- fct_relevel(.$origine, "01"),
    
    # 1 corresponds to being a man
    homme <- fct_relevel(.$homme, "1"),
    
    # 1 corresponds to rural municipality
    tuu_r <- fct_relevel(.$tuu_r, "1")
  )

```

## b) Interpretation of orgine's coefficients

```{r level regression}

level_dependent_variable <- "salhoraire"

# reformulate is use to transform vector into official R formula
# reformulate(c("var1", "var2"), response = "target") returns
# target ~ var1 + var2
reg_formula <- reformulate(termlabels= interest_variables, response = level_dependent_variable)

level_reg <- lm(reg_formula, data = df_last_wage)

summary(level_reg)

```

As said before, factored variables can be interpreted relatively to the comparator modality. Thus the regression seems to work because we found that the more you lived in a big city (higher factor tuu_r) the more the model predicts you a higher salary (+ 2 euros if you live in Paris conurbation). Factor(homme) being positive makes the model predicts a higher hourly salary if you are a man than a woman (+ 2 euros). Both of these variables seems to be relevant as the p_value suggest it (statistically significant at the 0.001% level).

Even if revised_age and revised_age squared are not statistically significant, them being positive fits our intuition. 

AS for the origin variable, we also have qualitatively quite good coefficients. Indeed almost every origin leads people to have a smaller hourly wage than a person having french origin at the exception of Northern Europe and Middle East origin. One may easily argue that the former one is due to a higher confindence granted to them than french origin workers. However, very few of them are statistically significant. The two main one are people with North African origin (06) or other African origin (07), perhaps mainly due to a larger diaspora of people of these origins in France. Both of them are negative, according to the model a person with African orgin receives all other things being equal about -2 euros from their hourly salary than french origin people. It is sadly what we could have anticipated, may be due to systemic racism in our society.



```{r log_linear regression}

log_dependent_variable <- "log(salhoraire)"

reg_formula <- reformulate(termlabels= interest_variables, response = log_dependent_variable)

log_reg <- lm(reg_formula, data = df_last_wage)

summary(log_reg)

```
As expected we obtain the same result than in the level regression, presented in a different form but with a much higher statistically significativity (especially for revised_age variable).

## c) Diffusion channels of inequalities due to the origin of people

To answer this question we use an iterative method. We first compute a simple regression of hourly salary on origin, then we do for each new potential channel variable a new regression where we add this variable to the vector of interest variables. 
By comparing the coefficient of origin in each of these regressions we are able to highlight some of the channels.

```{r regression over origin}

log_reg_origin <- lm(log(salhoraire) ~ factor(origine), data= df_last_wage)

summary(log_reg_origin)

```

```{r regression over origin and education}

log_reg_education <- lm(log(salhoraire) ~ factor(origine) + factor(education), data= df_last_wage)

summary(log_reg_education)

```


```{r regression over orign, education and contrat}

log_reg_contract <- lm(log(salhoraire) ~ factor(origine) + factor(education) + factor(contrat), data= df_last_wage)

summary(log_reg_contract)

```
```{r regression over origin, education, type of contract and health}

log_reg_health <- lm(log(salhoraire) ~ factor(origine) + factor(education) + factor(contrat) + factor(sante_declaree), data= df_last_wage)

summary(log_reg_health)

```

For comparing the evolution of origin's coefficients we use the library stargazer which helps to process the information in more readable format.

```{r evolution of coefficients of origin}

stargazer(log_reg_origin, log_reg_education, log_reg_contract, log_reg_health, 
          type = "text", 
          column.labels = c("Origin", "+Edu", "+Contract", "+Health"))

```


As we can see, adding education to the regression highly modify origin's coefficient. For instance, if we stick to the most significant coefficient which correspond to people with African origin (expect northern Africa) - modality 7 - we see that it increases following the addition of education as an interest variable. Without, the model predicts that these former people receive an hourly wage 17 % smaller than someone with french origin, whereas it predicts only 15 % smaller if we add education. Thus education explains a part of wage inequalities which means that the wage differences is partially ($(0.178 - 0.157) / 0.178 \approx 11.8\%$) conveyed through the education channel. The origin coefficient still remains significant which shows that other diffusion channels exists.

When the coefficient decreases following the insertion of education, which is only the case for factor 5 - Eastern european , this shows that the people with this origin are not paid enough for their level of education. Indeed, if the coefficient is initially close to 0 in the first regression the model predicts for these people the same hourly wage than for the reference group. By adding the education, we see that the model predicts a smaller wage all else equals but education. This mean that the model tend to predict that this person have in general a higher education level than what they are paid for.

Similarly, adding the type of contract increase all orgin's coefficient (except for factor 9 corresponding to Laotian, Cambodian, Vietnamese). The type of contract explains about $(refaire le calcule pour facteur 3 11.8\%$ of discrimination in hourly wage.

Finally, adding personal assessment of an individual's health doesn't modify the coefficient. Therefore origin of people doesn't seem to affect the hourly wage through the health of people. Since health status was not correlated with origin in our sample, it cannot explain why wages differ by origin.

# Question 2







## 











Adding control variables to resolve partially the endegoneous problem

Our goal is here to obtain a causal regression all else equals to interpretate wage differences as inequalities. But we face usual endogeneic problems in econometric. Endogeneity appears when E[Xepsilon] =/ 0, that is to say when interest variables are correlated with the residuals. This is all-ways the case because it is obviously impossible to explain the effect of a specific variable link to the real world with a finite number of interest variables, but we can try to have an expectation near 0. 

This endogeneity issue is not a problem as long as we only try to predict a result according to a regression model. But here we want to have a causal interpretation of coefficients of ours regressions. To do that we need to erase all bias due to omitted variables.

However we need to be really cautious to not include control variables that are a consequence of the origin of people. In this case the causal effect of origin will be partially captured in the control variable and the inequalities will seems smaller that what they really are (risk of bad controls)

```{r education variable}

# we add education to the vector of interest variables
interest_variables_educ <- c(interest_variables, "factor(education)")

# releveling "education" variable for having the right comparison in the regression
df_last_wage <- df_last_wage %>% 
  mutate(
    # 1 corresponds to a person that has obtain a degree above two years post-baccalaureate
    education <- fct_relevel(.$education, "1")
    )

```


```{r level regression with education}

reg_formula <- reformulate(termlabels= interest_variables_educ, response = level_dependent_variable)
level_educ_reg <- lm(formula= reg_formula, data= df_last_wage)

summary(level_educ_reg)

```


# Part 2




