---
title: "R_DM_2"
output:
  pdf_document: default
  html_document: default
---

Library importation
regarder stargazer si le temps

```{r necessary library}
library(haven)
library(dplyr)
library(forcats)
library(stargazer)
library(fixest)
library(plm)
```


# Data visualisation

```{r data importation}
file = "data/DM_Subject_2_Data.dta"
df <- read_dta(file)

# correction of the salmensuel's computing 
df$salhoraire <- df$salmensuel / (df$heuretra * 4.33)

head(df)
```

# Part 1


In this part, we looked at employed people, thus we only keep observations in df with a wage associated 

```{r filtering data by hourly wage}

df_wage <- df %>% 
  filter(!is.na(salhoraire))

```


`r nrow(df_wage)` observations over `r nrow(df)` have a mensual salary indicated. Indeed the wage is only available in the first and last quarter of the dataframe for employed individuals.



# Question 1 

We ignore the panel dimension in this question by keeping only the last observed quarter for employed individuals.

```{r filtering data by timestamp}

df_last_wage <- df_wage %>% 
  filter(t == 6)

```


Creation of new explanatory variables used in our linear regression.
Revised_age is a variable where we rescale age in order to have a quadratic relationship between the hourly salary and the age of the worker.

```{r creation of explanatory variables}

df_last_wage$revised_age <- df_last_wage$age - min(df_last_wage$age)

# capturing the quadratic relation
df_last_wage$revised_age_squared <- df_last_wage$revised_age ^ 2

```


## a) Linear regression results


For explanatory variables we obviously want to include origine because that's what explanatory us in the regression. To ensure capturing the real prediction of origine on the hourly wage, we need to include other variables that let unchanged the hourly salary all other things being equal. These variables should be related to the hourly wage but not with the origine. We thus selected 4 variables : 

  - "homme", an indicator of being a man (0 if a woman, 1 else)
  - "revised_age" and "revised_age_squared" (as described above)

These first two variables can quite likely be decorrelated of the origine of people

Lastly we hypothesize that the size of the urban area where people lived are not correlated to there origin which could be debate because rural area are often inhabted by french origin persons.

```{r explanatory variables}

# as_factor used for nominal variables
explanatory_variables <- c("factor(origine)", "factor(homme)", "revised_age", "revised_age_squared", "factor(tuu_r)") 

# as_factor takes as an entry a variable under character or factor type
df_last_wage$homme <- as.character(df_last_wage$homme)

# as_factor interpret coefficients relatively to the first level in alphebitc order. To have an easier interpretation of the coefficient we will ensure that the compared modality will be France (01) for origin, Man (1) for homme and rural municipality for tuu_r (1)

df_last_wage <- df_last_wage %>% 
  mutate(
    # 01 corresponds to a french origin
    origine = fct_relevel(origine, "01"),
    
    # 1 corresponds to being a man
    homme = fct_relevel(homme, "1"),
    
    # 1 corresponds to rural municipality
    tuu_r  = fct_relevel(tuu_r, "1")
  )

```

## b) Interpretation of orgine's coefficients

```{r level regression}

level_dependent_variable <- "salhoraire"

# reformulate is use to transform vector into official R formula
# reformulate(c("var1", "var2"), response = "target") returns
# target ~ var1 + var2
reg_formula <- reformulate(termlabels= explanatory_variables, response = level_dependent_variable)

level_reg <- lm(reg_formula, data = df_last_wage)

summary(level_reg)

```

As said before, factored variables can be interpreted relatively to the comparator modality. Thus the regression seems to work because we found that the more you lived in a big city (higher factor tuu_r) the more the model predicts you a higher salary (+ 2 euros if you live in Paris conurbation). Factor(homme) being positive makes the model predicts a higher hourly salary if you are a man than a woman (+ 2 euros). Both of these variables seems to be relevant as the p_value suggest it (statistically significant at the 0.001% level).

Even if revised_age and revised_age squared are not statistically significant, them being positive fits our intuition. 

AS for the origin variable, we also have qualitatively quite good coefficients. Indeed almost every origin leads people to have a smaller hourly wage than a person having french origin at the exception of Northern Europe and Middle East origin. One may easily argue that the former one is due to a higher confindence granted to them than french origin workers. However, very few of them are statistically significant. The two main one are people with North African origin (06) or other African origin (07), perhaps mainly due to a larger diaspora of people of these origins in France. Both of them are negative, according to the model a person with African orgin receives all other things being equal about -2 euros from their hourly salary than french origin people. It is sadly what we could have anticipated, may be due to systemic racism in our society.



```{r log_linear regression}

log_dependent_variable <- "log(salhoraire)"

reg_formula <- reformulate(termlabels= explanatory_variables, response = log_dependent_variable)

log_reg <- lm(reg_formula, data = df_last_wage)

summary(log_reg)

```
As expected we obtain the same result than in the level regression, presented in a different form but with a much higher statistically significativity (especially for revised_age variable).

## c) Diffusion channels of inequalities due to the origin of people

To answer this question we use an iterative method. We first compute a simple regression of hourly salary on origin, then we do for each new potential channel variable a new regression where we add this variable to the vector of explanatory variables. 
By comparing the coefficient of origin in each of these regressions we are able to highlight some of the channels.

```{r regression over origin}

log_reg_origin <- lm(log(salhoraire) ~ factor(origine), data= df_last_wage)

summary(log_reg_origin)

```

```{r regression over origin and education}

log_reg_education <- lm(log(salhoraire) ~ factor(origine) + factor(education), data= df_last_wage)

summary(log_reg_education)

```


```{r regression over orign, education and contrat}

log_reg_contract <- lm(log(salhoraire) ~ factor(origine) + factor(education) + factor(contrat), data= df_last_wage)

summary(log_reg_contract)

```
```{r regression over origin, education, type of contract and health}

log_reg_health <- lm(log(salhoraire) ~ factor(origine) + factor(education) + factor(contrat) + factor(sante_declaree), data= df_last_wage)

summary(log_reg_health)

```

For comparing the evolution of origin's coefficients we use the library stargazer which helps to process the information in more readable format.

```{r evolution of coefficients of origin}

stargazer(log_reg_origin, log_reg_education, log_reg_contract, log_reg_health, 
          type = "text", 
          column.labels = c("Origin", "+Edu", "+Contract", "+Health"))

```


As we can see, adding education to the regression highly modify origin's coefficient. For instance, if we stick to the most significant coefficient which correspond to people with African origin (expect northern Africa) - modality 7 - we see that it increases following the addition of education as an explanatory variable. Without, the model predicts that these former people receive an hourly wage 17 % smaller than someone with french origin, whereas it predicts only 15 % smaller if we add education. Thus education explains a part of wage inequalities which means that the wage differences is partially ($(0.178 - 0.157) / 0.178 \approx 11.8\%$) conveyed through the education channel. The origin coefficient still remains significant which shows that other diffusion channels exists.

When the coefficient decreases following the insertion of education, which is only the case for factor 5 - Eastern european , this shows that the people with this origin are not paid enough for their level of education. Indeed, if the coefficient is initially close to 0 in the first regression the model predicts for these people the same hourly wage than for the reference group. By adding the education, we see that the model predicts a smaller wage all else equals but education. This mean that the model tend to predict that this person have in general a higher education level than what they are paid for.

Similarly, adding the type of contract increase all orgin's coefficient (except for factor 9 corresponding to Laotian, Cambodian, Vietnamese). The type of contract explains about $(refaire le calcule pour facteur 3 11.8\%$ of discrimination in hourly wage.

Finally, adding personal assessment of an individual's health doesn't modify the coefficient. Therefore origin of people doesn't seem to affect the hourly wage through the health of people. Since health status was not correlated with origin in our sample, it cannot explain why wages differ by origin.

# Question 2

##  a) Education : a bad control variable

Our goal is here to obtain a causal regression all else equals to interpret wage differences as inequalities. But we face usual endogeneic problems in econometric. Endogeneity appears when E[Xepsilon] =/ 0, that is to say when explanatory variables are correlated with the residuals. This is always the case because it is obviously impossible to explain the effect of a specific variable link to the real world with a finite number of explanatory variables, but we can try to have an expectation near 0. 

This endogeneity issue is not a problem as long as we only try to predict a result according to a regression model. But here we want to have a causal interpretation of coefficients of our regressions. To do that we need to erase all bias due to omitted variables.

It is easy to support that education plays a role in hourly wage differences. But it is well known in economic that the level of education of people depends on a lot of different factors impossible to capture (motivation, talent, ...). In order to still have a causal interpretation of origin's coefficient we use an instrumental variable to isolate the role of education towards wage inequalities which is necessary to have a causal interpretation of all other explanatory variables. 

Moreover, we need to be really cautious when we include control variables that are a consequence of the origin of people because we are in presence of a post-treatment bias (or mediation bias). In this case the causal indirect effect of origin will be partially captured in the control variable and the inequalities will seem smaller that what they really are (risk of bad controls). 
Education falls in this category, but it being nonetheless one of the major channel where inequalities go through,  it is useful to try to add it as a explanatory variable. By adding it we will in the rest of the question focus on the direct role of origin in wage inequalities. The model without education measures the total systemic disadvantage starting from birth, whereas the model with education focuses on direct labor market discrimination. By controlling for degrees, we can observe the wage gap that exists between individuals who possess equal skills and schooling.
To have the total effect of origin over hourly wage differences one can look at the coefficient of the simple regression on question 1/c) where we regress 'salhoraire' on 'origine'.


## b) Choice of instrumental variables

The purpose of instrumental variables is to isolate the part of the education level that is independent of unobservable factors, such as innate talent or motivation. By doing so, we ensure that the estimated coefficient represents the true causal effect of education on wages, rather than reflecting these omitted variables. To do that, an instrumental variable should be :
  - relevant = be correlated to the explanatory variable (cov(Z,X) /= 0)
  - exogenous = only influence the dependent variable through the explanatory variable (corr(Z,epsilon) = 0)
  
Besides to compute the model, instrumental variables need to respect an order condition. The number of instrumental variables (m) should be equal or superior to the number of explanatory endogenous variables (k). Here we assume that only education is endogenous, thus we need 1 or more instrumental variable. Having more than k instrumental variables allow to use a Sargan test to test the validity of our model (over identified model).

In this study, we propose to use the socioeconomic status (CSP) of both parents as instruments for the education level. While they are significantly correlated, including both allows for an over-identification test (Sargan test) to assess the validity of our instrumental block. Although parental CSP might be criticized for potential direct links to job networks, it remains the most robust predictor of educational attainment available in the dataset."


## c) Instrumental regression

```{r Two-stage least squared}

# leveling education to ensure the right factor as referent
df_last_wage <- df_last_wage %>% 
  mutate(
    # 1 corresponds to people having a degree above two years post-baccalaureate
    education = fct_relevel(education, "1")
  )

# classic formula for R format, formula for the second stage
second_stage_log_reg_formula <- reformulate(termlabels= explanatory_variables, response = log_dependent_variable)
second_stage_log_reg_formula <- paste(format(second_stage_log_reg_formula), collapse ="")

# first stage formula
first_stage_log_reg_formula <- "| factor(education) ~ factor(csp_mere) + factor(csp_pere)"

# we paste these two formula into one to fit the format of feols function of the package fixest.
log_reg_formula_iv <- as.formula(paste(second_stage_log_reg_formula, first_stage_log_reg_formula))


# 2SLS with fixest package to obtain the right standard error in the model prediction
log_reg_iv <- feols(log_reg_formula_iv, data = df_last_wage)

summary(log_reg_iv)

```


The two-stage least squared regression made thanks to the function feols of the fixest package comes with 3 different tests that allow us to see if we implement valid instrumental variable. 

The F-test (i.e first test) verify if our instrument variable are really relevant : correlated to the explanatory variable which is here education. It is a multilevel test where the null hypothesis is that the coefficient of each instrumental variables is null, H0 = betacsp_mere = betacsp_pere = 0. We search to reject H0. In this regression we reject all null hypothesis at 10% and event at 1% for most of the factor in education.
However, to have a good instrumental regression, it is customary to search to obtain a t-statistic for the F-test superior to 10, else coefficients of are instrumental variable are consider too small to well predict the education variable. The problem here is probably due to the amount of different variable becuause we regress over 62 degree of liberty : each socioeconomic status is divided into 'r length(unique(df_last_wage$csp_mere))' variable which each correspond to one modality because of the factoring process. 

The Sargan test null hypothesis is that all are instrument are valid. Therefore we don't want to reject this hypothesis. Having a p-value over 0.1 is enough to not being able to reject H0.

Fianlly, we need to see if this approach is worth it or no. OLS model is by definition more precise than a two-stage least square (smaller standard error). That's why we want to use the instrumental approach only if education is an endogenous variable in the OLS. For testing that we look at the Wu-Hausman test. This test compare the coefficient of education in the OLS and in the instrumental regression. If both coefficient are close enough one to another we concludes that it was not necessary to implement instrument variables. The null hypothesis of this test is that $beta_iv - beta_OLS = 0$. We here reject at a 1% level which shows that education was an edogenous in the OLS regression.

Technically, the model confirms the necessity of an instrumental variable approach (significant Wu-Hausman test) and the consistency of the instruments (non-significant Sargan test). However, the instruments are weak ($F < 10$). The high number of socioeconomic status (SES) categories—resulting in 62 degrees of freedom—dilutes the first-stage predictive power. Because of the weakness of the coefficient, the model might over fit the noise. Indeed by multiplying the modality we cluster are dataset into very small group which tend to let too much weight on outliers. 


To stabilize the results and achieve a reliable causal estimation, grouping these categories into broader clusters is essential. In order to do that we will regroup the socioeconomic status into the 7 reference group provide by INSEE : https://www.insee.fr/fr/metadonnees/pcs2020/groupeSocioprofessionnel/1?champRecherche=true

```{r instrumenatal regression with socioeconomic reference group}

# each reference group correspond to the first figure of the detailed group number 
df_last_wage <- df_last_wage %>% 
  mutate(
    reference_csp_mere = substr(csp_mere,1,1),
    reference_csp_pere = substr(csp_pere,1,1)
  ) 


# classic formula for R format, formula for the second stage
second_stage_log_reg_formula <- reformulate(termlabels= explanatory_variables, response = log_dependent_variable)
second_stage_log_reg_formula <- paste(format(second_stage_log_reg_formula), collapse ="")

# first stage formula
first_stage_log_reg_formula <- "| factor(education) ~ factor(reference_csp_mere) + factor(reference_csp_pere)"

# we paste these two formula into one to fit the format of feols function of the package fixest.
log_reg_formula_iv <- as.formula(paste(second_stage_log_reg_formula, first_stage_log_reg_formula))

# 2SLS with fixest package to obtain the right standard error in the model prediction
log_reg_iv <- feols(log_reg_formula_iv, data = df_last_wage)

summary(log_reg_iv)

```
Unlike the former regression, the the first stage F-statistic now exceed the rule_of_thumb threshold of 10. As fot the instrumental regression with detailed socioeconomic group, the p-values of F-tests and Sargan test shows that it this model is statistically valid.

However, it is now only possible to reject the Wu-Hausman test with a 10% significativity rate. Even if this evidence of endogeneity is more moderate than in the previous model, we will assume, because of the vast research on this topic in econometrics that education is an endogenous variable and should be treated as such. That's why this instrument variable is in our opinion better than an OLS on the log of the hourly wage over education and other explanatory variables.   


## d) Interpretation of origin's coefficients 

Thanks to the work done in previous questions, we are know able to interprate in a causal way the coefficient of the second instrumental regression. In the OLS model omitting education, both the 'Maghreb' (06) and 'Rest of Africa' (07) categories exhibited significant wage penalties of 10.3% and 22.1%, respectively (question 1/c). Once education is properly instrumented in our instrumental variable model, the coefficient for the Maghreb becomes statistically insignificant. This reversal suggests that for individuals of Northern African origin, the observed wage gap is primarily explained by differences in level of education  rather than direct labor market discrimination due to origin. 
In contrast, the penalty for the 'Rest of Africa' remains nearly identical and highly significant (approx. 22%) in both models. This indicates a persistent 'residual' disadvantage for this group that schooling alone cannot bridge. 


Furthermore, the gender pay gap widens from 13.6% in the OLS to 17.3% in the instrumental variable model, suggesting that the initial OLS was underestimating the true extent of gender inequality.

These findings confirm the necessity of our IV approach to accurately unpack the complex interactions between social origin, schooling, and earnings.


# Question 3


From now on, we use the whole dataset at our disposal to be able to erase all unobserved individual terms - unchanged in time - that bias coefficients with whom we are triying to have a causal interpretation of the hourly wage.

Panel regression takes the following form : $$y_{it} = \beta x_{it} + \alpha_i + \epsilon_{it}$$
Where :
  - $y_{it}$ is the dependent variable of individual i at time t (in this case the hourly wage)
  - $x_{it}$ are all explanatory variables of individual i at time t
  - $\alpha_i$ is the unobserved terms of individual i. these are invariant in time and represent effect such as his talent, his motivation, and so on. 
  - $\epsilon_{it}$ being the noise of the model


```{r panel dataframe for hourly wage}

df_panel_wage <- df_wage %>% 
  filter(t == 1 | t == 6)

# We transform the dataframe into a panel structure one required for using the plm package
# index allows to identify the person and the year
pdata_wage <- pdata.frame(df_panel_wage, index = c("id_individu", "t"))

```

We pre-process this dataframe we the same transformation that we have done in the first two questions

```{r panal dataframe preprocessing}


pdata_wage$revised_age <- pdata_wage$age - min(pdata_wage$age)

# capturing the quadratic relation
pdata_wage$revised_age_squared <- pdata_wage$revised_age ^ 2


pdata_wage$homme <- as.character(pdata_wage$homme)

# as_factor interpret coefficients relatively to the first level in alphebitc order. To have an easier interpretation of the coefficient we will ensure that the compared modality will be France (01) for origin, Man (1) for homme and rural municipality for tuu_r (1)

pdata_wage <- pdata_wage %>% 
  mutate(
    # 01 corresponds to a french origin
    origine = fct_relevel(origine, "01"),
    
    # 1 corresponds to being a man
    homme = fct_relevel(homme, "1"),
    
    # 1 corresponds to rural municipality
    tuu_r  = fct_relevel(tuu_r, "1"),
    
    # 1 corresponds to people having a degree above two years post-baccalaureate
    education = fct_relevel(education, "1")
  )


```


## a) Random effect panel regression

The goal of panel regression is therefore to obtain a causal interpretation of the effect of explanatory variables on the dependent variable free of all bias due to unobserved ability of individuals (motivation, ...).

Because these unobserved terms are time invariant, the most robust model who assure to have a valid causal interpretation is obtain by doing the usual least squared estimation of coefficient on a model where we substracts the mean over time from the original model.
This method, called the Fixed effect is super robust but very inflexible, indeed it unfortunatly also prevents to capture the effect of time invariant explanatory variables. 

Step 1 : Original model
$$y_{it} = \beta x_{it} + \alpha_i + \epsilon_{it}$$

Step 2 : new model obtained by de-meaning the first equation (Within transformation)
$$(y_{it} - \bar{y}_i) = \beta (x_{it} - \bar{x}_i) + (\alpha_i - \alpha_i) + (\epsilon_{it} - \bar{\epsilon}_i) \\ \iff \\ y^*_{it} = \beta x^*_{it} + \epsilon^*_{it}$$
Because for all time invariant terms $\frac{1}{n} \sum_{t=1}^{T} x_{it} = x_{it}$, we see there that they also disappear in this new equation.


Step 3 : estimation of the coefficients with the transformed equation (OLS)
$$\min_{\beta} \sum_{i=1}^{n} \sum_{t=1}^{T} (y^*_{it} - \beta x^*_{it})^2$$

Step 4 : Prediction 
$$\hat{y}_{it} = \hat{\alpha}_i + \hat{\beta} x_{it}$$
Where $\hat{\alpha}_i = \bar{y}_i - \hat{\beta} \bar{x}_i$


As said before the issue with this process is that we loose all time invariant explanatory variable which is here a major issue since we want to find inequalities in wage due to the origin of people, which obviously fall in the forgotten variable. To adress this issue an other process is possible : the random effect regression.

Instead of substracting the entire mean of variables with whom we work we only substracts a certain percentage ($\theta$) of it.

Thus, the first step remains the same, but we apply in step 2 a quasi-difference transformation :
$$(y_{it} - \theta\bar{y}_i) = (1 - \theta) \beta_{0} + \beta (x_{it} - \theta\bar{x}_i) +  \gamma (1 - \theta) z_{i}+(1 - \theta) \alpha_i + (\epsilon_{it} - \theta\bar{\epsilon}_i) \\ \iff \\ y^*_{it} = \beta x^*_{it} + \gamma z^*_{i} + \nu^*_{it}$$
where $\theta \in [0,1]$, $z_{i}$ are the time invariant explanatory variables and $\nu^*_{it} = (1 - \theta) \alpha_i + (\epsilon_{it} - \theta\bar{\epsilon}_i) $

Step 3 : estimation of the coefficients with the transformed equation (OLS)
$$\min_{\beta} \sum_{i=1}^{n} \sum_{t=1}^{T} (y^*_{it} - \beta x^*_{it} - \gamma z^*_{i})^2$$

In step 4, the prediction becomes :
$$\hat{y}_{it} = \hat{\beta}_0 + \hat{\beta} x_{it} + \gamma\hat{z}_{i}$$

For being able to have a causal interpretation of this model, the sine qua non condition is that the unobserved individual term are uncorrelated to all the explanatory variable : $Cov(X_{it}, \alpha_{i}) = 0, \forall i,t$ and $Cov(Z_{it}, \alpha_{i}) = 0, \forall i, t$.


Aside : finding the right $\theta$

$\theta$ is chosen so that it minimizes the serial correlation of errors within the same individual. 
$$\theta = 1 - \sqrt{\frac{\sigma^2_\epsilon}{\sigma^2_\epsilon + T \sigma^2_\alpha}}$$
Where T is the number of period in the panel dataframe

We first estimate the variance of the random noise by doing a fixed effects regression and then we estimate the one of the unobserved terms thanks to a between regression (difference over the mean of each person).


In this question we assume the exogeneity of the $\alpha_i$ and so we can use the random effects framework presented above. Because of exogeneity we can have a causal integration that includes invariant time explanatory variables in particular origin.

```{r random effects regression}

re_explanatory_variables <- c(explanatory_variables, "factor(education)")

re_log_formula <- reformulate(termlabels = re_explanatory_variables, response = log_dependent_variable)
print(re_log_formula)

re_log_reg <- plm(formula = re_log_formula, data =pdata_wage, model = "random")

```


Before being able to interpret the result of random effects regression we first need to test the exogeneity of the unobserved coefficients with an Hausman test. As the one in question 2, the Hausman test compared a robust model with a more precised one but potentially biased. If coefficients of the risked model corresponds to the one of the robust model (here the fixed effects), then it means the model is non biased. Because in fixed effects regression time invariant variables disappear, the null hypothesis of the Hausman test is : $H_0 : \hat\beta_{tvf} - \hat\beta_{tvr} = 0$ where tvf and tvr stands for respectively time variant variables in fixed effects and random effects model.


```{r fixed effects regression}

# fixed effects regression (random effects for theta = 1)
fe_log_reg <- plm (formula = re_log_formula, data = pdata_wage, model = "within")

# Hausman test
phtest(fe_log_reg, re_log_reg)

```
As we can see, we reject very significatively the null hypothesis of the Hausman test which mean that our random effects model is not consistant. The causal interpretation of coefficients that results from this model is thus impossible because they may incorporate the effect of the unobserved individual effects in it. Therefore the assumption according to which $ Cov(Z_{it}, \alpha_{i}) = 0, \forall i, t$ can statistically consider as being false. 

That being said, because we assume, wrongly, in the question that the $alpha_i$ are exogeneous to the explanatory variables, we still present below the result of the random effects model but these coefficients should not be seen as valid.

```{r unvalid results of random effects }

summary(re_log_reg)

```
As explains before $\theta$ depends on the number of period T, here we can see that it changes because we have an unbalanced panel, i.e that for some person we don't have both values of their hourly wage (for t = 1 and t = 6)

Around $\frac{2}{3}$ of the unexplained variance comes from the individual effects. Individual heterogeneity predominates which shows the importance to do a panel approach in this study.

## b) Origin variable in random effects model

This model can handle time invariant explanatory variables (cf previous question).
However, this model does not resolve potential endogeneity issues. On the contrary, it requires the strong assumption of exogeneity of individual effects ($Cov(X_{it}, \alpha_i) = 0$, \forall i, t) to be valid. If unobserved heterogeneity (such as motivation) is correlated with origin or education, the estimator will be biased. In our case, since the Hausman test rejected the null hypothesis, we know that this model fails to control for this endogeneity bias.


## c) correlated unobserved heterogeneity

In question 3/ a) we saw that a fixed effects model is just a specific case of random effects model where $\theta = 1$. If $\theta = 1$ then all time invariant variables, such as origin, disappear and can't be handle by the model.



# Part 2


In this part, we focus on the effect of the origin on the fact of being employed if unemployed.

```{r data preprocessing}

# acteu = 1 if the person is employed, 2 if unemplyed and 3 if inactive
df_active <- df %>% 
  filter(acteu %in% c(1,2))

# We transform the dataframe into a panel structure one required for using the plm package
# index allows to identify the person and the year
pdata_active <- pdata.frame(df_active, index = c("id_individu", "t"))

```

`Acteu` is the binary dependent variable we want to study, but it's more convenient to take a binary variable that takes its values in {0,1} which is better for having a probability interpretation of the results. We create from `chomeur` (indicator of being unemployed), the indicator of being employed which we will use in the next question as the dependent variable. The closest the predicted value will be closed to one, the more the model will predict a high probability that the individual will be employed.

```{r indicator of being emplyed}

pdata_active <- pdata_active %>% 
  mutate(
    being_employed = ifelse(chomeur == 1, 0, 1)
  )

```


# Question 4

Having a binary dependent variable raises issues regarding interpretation of the coefficients if we use a linear estimation. Indeed we can obtain for extreme values of our explanatory variables a predicted value above 1 or under 0 which has no sens considering this is a probability of being employed. But if we are primarily interested in the average marginal effect (the slope of the regression line) rather than individual prediction then an OLS or panel model remains a robust and widely accepted approach in empirical economics. 

## a) random effects model


```{r regression variables}

binary_dependent_variable <- "being_employed"
explanatory_variables <- c("factor(origine)", "factor(homme)", "revised_age", "revised_age_squared", "factor(tuu_r)", "factor(education)")



pdata_active$revised_age <- pdata_active$age - min(pdata_active$age)

# capturing the quadratic relation
pdata_active$revised_age_squared <- pdata_active$revised_age ^ 2


pdata_active$homme <- as.character(pdata_active$homme)

pdata_active <- pdata_active %>% 
  mutate(
    # 01 corresponds to a french origin
    origine = fct_relevel(origine, "01"),
    
    # 1 corresponds to being a man
    homme = fct_relevel(homme, "1"),
    
    # 1 corresponds to rural municipality
    tuu_r  = fct_relevel(tuu_r, "1"),
    
    # 1 corresponds to people having a degree above two years post-baccalaureate
    education = fct_relevel(education, "1")
  )

```


In this question we use the exact same processus than in question 3/

```{r random effect }

re_binary_formula <- reformulate(termlabels = explanatory_variables, response = binary_dependent_variable)
print(re_binary_formula)

re_binary_reg <- plm(formula = re_binary_formula, data =pdata_active, model = "random")

```

```{r}
summary(re_binary_reg)
```






